**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**

### 2015-10-01 Schloss_Cluster_PeerJ_2015.Rmd
Working on R&D section

```{r, cache=TRUE}
n_miseq_seqs <- length(scan("data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.redundant.fix.fasta", what="", sep="\n", quiet=T))/2
```

***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering algorithms. First, the Canadian soil dataset that the He study used to analyze the stability of OTUs is one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform. Developments in sequencing technology now permits the sequencing of millions of sequences for a study. In addition, because the original phred quality scores and flowgram data are not available, it was not possible to adequately remove sequencing errors (REFS). The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering algorithms. Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene. For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference algorithms. Of the XXXXXX sequences in the default QIIME database, only XXXXXX fully cover the V9 region. Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene. With these points in mind, we compared the clustering assignment for each of these algorithms using a time series experiment that was obtained using mouse stool (REFS). The MiSeq platform was used to generate `r ` sequences from the V4 region of the 16S rRNA gene of XXX samples. Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02%. Although no dataset is perfect for exhaustively testing these clustering algorithms, this dataset was useful for demonstrating several points.

@@

@@

***Additional problems with reference-based clustering.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs. This was unsurprising since, by definition, the algorithm represents a one-to-one mapping of reads to a reference and treats the input sequences independently. An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database. We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference algorithm with the full MiSeq dataset. Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations. On average there were `r round(rand_ref_u["mean", 1], 2)` OTUs per randomization (range from `r round(rand_ref_u["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`). The unrandomized reference resulted in `r rand_ref_u["original", 1]` OTUs, below the minimum observed number of OTUs when the references were randomized. This surprising result is likely because a heuristic is employed by USEARCH to determine when to stop searching the database such that the best possible match is not necessarily returned. If the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points. To test this, we substituted VSEARCH for USEARCH in the closed-reference algorithm because VSEARCH does not utilize the heuristic but is otherwise modeled after USEARCH. In fact, all randomizations of the reference sequences and the original ordering of the reference sequences generated `r rand_ref_v["original", 1]` OTUs.

@@

v19_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)
n_v19_good <- sum(v19_summary$start <= 3967 & v19_summary$end >= 6116)
n_v4_good <- length(scan("data/gg_13_8/gg_13_8_97.v4_ref.names", what="", sep="\n", quiet=T))

v_overlap <- read.table(file="data/rand_ref/overlap.vclosed.summary")
```

We noticed that even with VSEARCH, the reference OTU label that was assigned to each OTU differed even though the composition of the OTUs were the same across the randomized references. The default QIIME database that was used in the He study and by others contains full-length sequences that are at most 97% similar to each other. We suspect that the reference database was designed this way to overcome the limitations of the heuristic employed in USEARCH. This curated database is used for mapping sequences to a reference phylogenetic tree for carrying out UniFrac-based analyses (REFS) and to reference genomes for performing analyses such as picrust (REFS). Because the downstream applications depend on the correct and unique labeling of the OTUs, the lack of stability of the labeling was a concern. We hypothesized that within a given region  there would be sequences that were more similar to each other and possibly identical to each other. There were a total of `r nrow(v19_summary)` sequences in the reference database and `r n_v19_good` of those sequences fully overlap the V4 region. Among those sequences that fully overlapped the V4 region, there were `r n_small_v19_dist` pairs of sequences that were more than 97% similar to each other over the full length of the 16S rRNA gene. When the extracted V4 sequences were dereplicated, we identified `r n_v19_good-n_v4_good` sequences that were identical to each other. Among the dereplicated V4 sequences there were `r n_small_v4_dist` pairs of sequences that were more than 97% similar to each other. The presence of duplicate V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference algorithm. Using VSEARCH to perform the closed-reference clustering, an average of `r round(100*v_overlap["orig", "mean"], 1)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 1)` to `r round(100*v_overlap["orig", "max"], 1)`).

```{r duplicate_analysis}
dups <- read.table(file="data/gg_13_8/duplicate.analysis")
max_taxa <- dups[dups$V3 == max(dups$V3),]
```

We next asked whether the lack of consistency in labeling would affect the taxonomic identification of the sequences. The `r sum(dups$V2)-nrow(dups)` duplicated V4 sequences represented `r nrow(dups)` unique sequences. Among the `r sum(dups$V2==2)` reference sequences that had one duplicate, `r sum(dups$V2==2 & dups$V3==2)` had discordant taxonomies. Among those `r sum(dups$V2 > 2)` sequences with two or more duplicates, `r sum(dups$V2>2 & dups$V3>=2)` had discordant taxonomies. Two sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies. There was also a sequence had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies.

***Evalution of an open-source alternative to USEARCH.***

>>>>

```{r, cache=TRUE}
gg_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)
n_v4_overlap <- sum(gg_summary$start <= 3967 & gg_summary$end >= 6116)
n_v9_overlap <- sum(gg_summary$start <= 10179 & gg_summary$end >= 10683)

n_miseq_seqs <- nrow(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.redundant.summary"))

n_unique_miseq_seqs <- nrow(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.summary"))

n_groups <- ncol(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.denovo.uchime.pick.pick.count_table")) - 2 #subtract 2 to remove the sequence name and total number of sequences
```

***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering algorithms. First, the Canadian soil dataset that the He study used to analyze the stability of OTUs is one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform. Developments in sequencing technology now permits the sequencing of millions of sequences for a study. In addition, because the original phred quality scores and flowgram data are not available, it was not possible to adequately remove sequencing errors (REFS). The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering algorithms. Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene. For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference algorithms. Of the `r nrow(gg_summary)` sequences in the default QIIME database, only `r n_v4_overlap` fully cover the V9 region. In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region. Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs. Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene. With these points in mind, we compared the clustering assignment for each of these algorithms using a time series experiment that was obtained using mouse stool (REFS). The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples. Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% (REFS). Although no dataset is perfect for exhaustively testing these clustering algorithms, this dataset was useful for demonstrating several points.

@@

***Evalution of an open-source alternative to USEARCH.*** The AGC and DGC algorithms appear to perform as well and possibly than the hierarchical clustering algorithms for some datasets. These algorithms utilize the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program. Access for non-academic users and those needing the 64-bit version is available commercially from the developer. An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as a open-source alternative. One subtle difference between the two programs is that USEARCH employs a heuristic to determine when to stop searching a database. This means that the best possible match is not necessarily returned once a highly similar hit is obtained. In contrast, VSEARCH does not utilize the heuristic, which the VSEARCH developers claim enhances the sensitivity relative to USEARCH. Using the two datasets, we determined whether the AGC and DGC algorithms as implemented by the two methods yielded sequencing assignments of similar quality.

 This also explains the variation observed in the AGC and DGC algorithms, which rely upon the USEARCH algorithm.
[   Something, something, something.    ]

@@

***Problems with reference-based clustering.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs. This was unsurprising since, by definition, the algorithm represents a one-to-one mapping of reads to a reference and treats the input sequences independently. An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database. The default QIIME database, which was used in the He study and by others, contains full-length sequences that are at most 97% similar to each other. We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference algorithm with the full MiSeq dataset, which contained `r n_unique_miseq_seqs` unique sequences. Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations. On average there were `r round(rand_ref_u["mean", 1], 2)` sequences that mapped to a reference OTU per randomization (range from `r round(rand_ref_u["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`). The original ordering of the reference resulted in `r rand_ref_u["original", 1]` sequences being mapped, less than the minimum observed number of mapped sequences when the references were randomized. This surprising result was likely due to the performance of the USEARCH heuristic. As with the AGC and DGC algorithms, if the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points. To test this further, we substituted VSEARCH for USEARCH in the closed-reference algorithm. When we used VSEARCH the original ordering of the reference sequences and all randomizations were able to map `r rand_ref_v["original", 1]` sequences to reference OTUs. That VSEARCH mapped more sequences than even the maximum number of sequences found across the USEARCH randomizations supports the VSEARCH developers' assertion that it has greater sensitivity than USEARCH.

@@

v19_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)
n_v4_overlap
n_v4_unique <- length(scan("data/gg_13_8/gg_13_8_97.v4_ref.names", what="", sep="\n", quiet=T))

v_overlap <- read.table(file="data/rand_ref/overlap.vclosed.summary")

dups <- read.table(file="data/gg_13_8/duplicate.analysis")
max_taxa <- dups[dups$V3 == max(dups$V3),]
```

We also observed that regardless of whether we used USEARCH or VSEARCH, the reference OTU labels that were assigned to each OTU differed between randomizations. When we used USEARCH to perform closed-reference clustering, an average of `r round(100*u_overlap["orig", "mean"], 1)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*u_overlap["orig", "min"], 1)` to `r round(100*u_overlap["orig", "max"], 1)`). If we instead used VSEARCH an average of `r round(100*v_overlap["orig", "mean"], 1)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 1)` to `r round(100*v_overlap["orig", "max"], 1)`). To better understand this result, we further analyzed the reference database that is used in QIIME. We hypothesized that within a given region there would be sequences that were more than 97% similar and possibly identical to each other. When a sequence was used to search the randomized databases, it would encounter a different reference sequence as the first match with each randomization. Among those reference sequences that fully overlap the V4 region, there were `r n_small_v19_dist` pairs of sequences that were more than 97% similar to each other over the full length of the 16S rRNA gene. When the extracted V4 sequences were dereplicated, we identified `r n_v4_unique` unique sequences. Among these dereplicated V4 sequences there were `r n_small_v4_dist` pairs of sequences that were more than 97% similar to each other. The presence of duplicate V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference algorithm. We suspect that the reference database was designed to only include sequences that were at most 97% similar to each other way to overcome the limitations of the USEARCH search heuristic.

Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference algorithms to generate OTU labels that are used in several downstream applications.  to extract information from a reference phylogenetic tree to carrying out UniFrac-based analyses (REFS) and to identify reference genomes for performing analyses such as PICRUSt (REFS). Because these downstream applications depend on the correct and unique labeling of the OTUs, the lack of stability of the labeling is problematic. As one illustration of the effects that incorrect labels would have on an analysis, we asked whether the duplicate sequences had the same taxonomies. Among the `r sum(dups$V2==2)` reference sequences that had one duplicate, `r sum(dups$V2==2 & dups$V3==2)` had discordant taxonomies. Furthermore, among those `r sum(dups$V2 > 2)` sequences with two or more duplicates, `r sum(dups$V2>2 & dups$V3>=2)` had discordant taxonomies. Two sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies. There was also a sequence had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies. Together, these results demonstrate some of the considerable problems with the reference-based clustering of sequences.

## Conclusions
* Stability is meaningless in context of accuracy and precision of clustering assignments
* *De novo* clustering algorithms outperform reference-based approaches.
* The best algorithm is data-specific and researchers should evaluate the clustering obtained by multiple methods

* Don't use USEARCH to cluster sequences
* VSEARCH is an excellent alternative

* Reference-based clustering has many problems

* Many difficulties in clustering have to do with underlying data quality.
